---
title: "Attractors"
author: "George Savva"
format:
  revealjs:
    theme: [dark]
    slide-number: true
    #css: "custom.css"
    scrollable: true
    fig-align: center
    echo: true
    transition: slide
    incremental: false
    self-contained: true
---


```{r echo=FALSE}
library(imager)
l=seq(-2,2,l=1000)
par(bg=hsv(0,0,0,0), mar=c(0,0,0,0))
```

# 1. Hubris

::: {.notes}
I started making art with the idea that computers ought to be better at making it than people

I had the idea that, after all, an image is just a set of pixels, and there's no reason why you couldn't searh through the possible arrangements of these pixels, optimising for some criterion 

This is an idea that'd been rumbling around in my mind for a while, but I didn't have the computing tools necessary to do this until about 2019... I'd picked up the idea and played with it for a while back in the 00s when I was a student but a career in proper statistics distracted me...

But working again in computational biology, the idea of evolving images towards some objective measure of beauty was appealing.


OK - so we need a representation of an image.

Note - I'm not talking about AI art or machine learning here.. what I'm dealing with is fashioned maths and programming.  I have no interest at all in black box mimicry, leaves me cold.  I need to see the entire process.

So...a statistical programmer with no training at all is going to invent a system to define an image in maths, and then steer the maths towards something cool.

Now, all this was going to rest on how you describe an image to a computer, and then how you manipulate this.


:::

## What is an image?

An image is defined by its colour at every point.  

So any mapping of locations to colours defines an image!

Locations are two dimensional, colours are three dimensional

So any function:

$$f: \mathbb{R}^2 \rightarrow \mathbb{R}^3$$

defines an image, and every image is such a function.

---  

$$
f(x,y)=\left\{\begin{array}{ll}
0 & \text{if}\quad x^2+y^2<1\\
1 & \text{otherwise}
\end{array}\right.
$$

```{r echo=FALSE}
#| code-line-numbers: 2
outer(X=l,Y=l,\(X,Y) 
      X^2+Y^2 > 1 
) |> as.cimg() |> plot(ax=F,as=1)
```

---  

For a fake Malevich:  

$$
\text{f(x,y)}=\left\{\begin{array}{ll}
0 & \text{if}\quad \max(|x|, |y|)<1.5\\
1 & \text{otherwise}
\end{array}\right.
$$


```{r}
#| code-line-numbers: 2
outer(X=l,Y=l,\(X,Y) 
      pmax(Mod(X),Mod(Y)) > 1.5 
) |> as.cimg() |> plot(ax=F,as=1)

```


---

$$
f(x,y)=(\sin(3x)^2+\cos(3y)^2)\ \text{Mod}\ 0.2 > 0.1
$$


```{r}
#| code-line-numbers: 2
outer(X=l,Y=l,\(X,Y) 
      (sin(X*3)^2 + cos(Y*3)^2)%%0.2 > 0.1
) |> as.cimg() |> plot(ax=F,as=1)

```

---

```{r echo=F}
plotfunctions <- function(h,s,v,X,Y){
    a <- array(dim=c(length(X),length(Y),3))
  a[,,1] <- 180*(Mod(outer(X,Y,h))%%1)
  a[,,2] <- outer(X,Y,s)%%1
  a[,,3] <- outer(X,Y,v)%%1
  #print(a)
  b=HSVtoRGB(as.cimg(a))
  #print(as.array(b))
    par(bg=hsv(0,0,0,0), mar=c(0,0,0,0))
  b |> plot(ax=F,as=1, rescale=F)
}
```

```{r echo=F}
plotfunctionsC <- function(h,s,v,X,Y){
    a <- array(dim=c(length(X),length(Y),3))
  a[,,1] <- 360*(Mod(outer(X,Y,\(X,Y) h(X+1i*Y)))%%1)
  a[,,2] <- outer(X,Y,\(X,Y) s(X+1i*Y))%%1
  a[,,3] <- outer(X,Y,\(X,Y) v(X+1i*Y))%%1
  #print(a)
  b=HSVtoRGB(as.cimg(a))
  #print(as.array(b))
  par(bg=hsv(0,0,0,0), mar=c(0,0,0,0))
  b |> plot(ax=F,as=1, rescale=F)
}


```

--- 

```{r echo=T}
plotfunctions(
  h=\(X,Y) X/2,
  s=\(X,Y) sin(3*X)+cos(3*Y),
  v=\(X,Y) cos(3*X)+sin(3*Y),
  l,l
  )
```

--- 

```{r echo=T}
plotfunctions(
  h=\(X,Y) X/2,
  s=\(X,Y) sin(3*X)+cos(3*Y)+.1*rnorm(length(X)),
  v=\(X,Y) cos(3*X)+sin(3*Y)+.1*rnorm(length(X)),
  l,l
  )
```

--- 

```{r echo=T}
plotfunctionsC(
  h=\(Z) Arg(atan(Z)),
  s=\(Z) Mod(atan(Z)),
  v=\(Z) Arg(Z*10)+Mod(Z),
  l,l
  )
```

--- 

# What makes an image good?



# 2. Humility

What I'd done is reinvent domain colouring. 

This is a mathematical technique that represents simple functions, but uses colours to represent values.  It's typically applied as a visualisation tool, people don't seem to have thought about it much from a purely aesthetic point of view.


# 3. History

# 4. Chaos

# 5. Victorian